---
title: "Seascape Ecology"
subtitle: "General notes"
author: "Ryan Reisinger"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Tracking data

## Filtering tracking data

Argos animal tracking data locations have different accuracies, indicated by the 'location class'. From the Argos page (<https://www.argos-system.org/support-and-help/faq-localisation-argos/>):

> -   classes 0, 1, 2, 3 indicate that the location was obtained with 4 messages or more and provides the accuracy estimation,
>
> -   class A indicates that the location was obtained with 3 messages,
>
> -   class B indicates that the location was obtained with 2 messages,
>
> -   class G indicates that the location is a GPS fix obtained by a GPS receiver attached to the platform. The accuracy is better than 100 meters.
>
> -   class Z indicates that the location process failed.
>
> The accuracy cannot be estimated for classes A and B (not enough messages).

See more in [Irvine et al.](https://animalbiotelemetry.biomedcentral.com/articles/10.1186/s40317-020-00207-x), [Costa et al.](https://doi.org/10.1371/journal.pone.0008677) and [Douglas et al.](https://doi.org/10.1111/j.2041-210X.2012.00245.x), for example.

We therefore need to filter these tracking data. We can use the `argosfilter` package (<https://cran.r-project.org/web/packages/argosfilter/index.html>) which has a simple speed filter (function `vmask()`) which excludes unrealistic location estimates based on their speed, and a speed-distance-angle filter (function `sdafilter()`), which takes into account the turning angle too.

You could also fit various state-space models, which fit a movement model to the tracking data and estimate likely locations, possibly at fixed time intervals (remember, Argos locations are not necessarily at fixed intervals, especially for diving animals).

Examples are Ian Jonsen's `bsam` (<https://cran.r-project.org/web/packages/bsam/index.html>) and `aniMotum` (<https://ianjonsen.github.io/aniMotum/>) packages, and Devin Johnson et al.'s `crawl` package (<https://cran.r-project.org/web/packages/crawl/index.html>). These are much more complicated to use than simple filtering, but can provide extra information like behavioural states.

# Colour palettes

The `viridis` package (<https://cran.r-project.org/web/packages/>) contains good colourblind-safe colour palettes that can be used with `ggplot2`. See the introduction at: <https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html>

```{r, eval = FALSE}
library(ggplot2)

?ggplot2::scale_colour_viridis_c()

ggplot2::ggplot()+
  scale_colour_viridis(limts=c(0,100),oob=scales::squish,values=c(0,0.25,0.4,1))
```

A huge variety of colour palettes from different packages and resources is indexed on this page:

<https://github.com/EmilHvitfeldt/r-color-palettes>
